---
title: "ANA 515 - Assignment 2 - Storm Event Analysis"
author: "Shanmugapriya Mohan Kumar"
date: "`r Sys.Date()`"
output:
  html_document:
    theme:
      bootswatch: minty
---
<style>
  h1.title {
    text-align: center;
  }
  h4.author {
    text-align: center;
  }
  h4.date {
    text-align: center;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, message = TRUE, warning = FALSE)
library(tidyverse)
library(knitr)
library(bslib)
library(readr)
library(rmarkdown)
library(dplyr)  # for data manipulation
library(lubridate)  # for date handling
library(stringr)
```

<style>
  h1.title {
    text-align: center;
  }
  h4.author {
    text-align: center;
  }
  h4.date {
    text-align: center;
  }
</style>

## Objective

The objective of this analysis is to process and visualize storm events data from a provided CSV file. 

1. Filter the storm events data to include only events listed by county FIPS.
2. Format the state and county FIPS codes and combine them into a single FIPS column.
3. Standardize all column names to lower case for consistency.
4. Integrate state-level information, including area and region, from built-in R datasets.
5. Calculate the number of storm events per state.
6. Merge the storm events data with state information, ensuring only matching states are included.
7. Display the merged data in a comprehensive table.
8. Create a scatter plot to visualize the relationship between land area and the number of storm events, with points color-coded by region.

Through this analysis, we aim to gain insights into the distribution and frequency of storm events across different states and regions, and how these events correlate with the geographic area of the states.

``` {r load-data, echo=FALSE}
# Dataset URL
stormevent_data <- "https://raw.githubusercontent.com/ShanmugapriyaMohankumar/Week6-Assignment/main/StormEvents_details.csv"

# Read the dataset from the URL
storm_data <- read_csv(stormevent_data,show_col_types = FALSE) #read_csv is part of the readr package, which is included in the tidyverse collection of R packages

#number of rows and columns
storm_data_rows<-nrow(storm_data)
storm_data_columns<-ncol(storm_data)

```
### Load and display data 
    The table contains `r storm_data_rows` rows and `r storm_data_columns` variables.

``` {r display-data, echo=FALSE}
#display the first few rows of the dataframe
head(storm_data)
```
# Limit the dataframe to the selected columns

``` {r limit-data, echo=FALSE}
# Select the desired columns
selected_columns <- c(
  'BEGIN_YEARMONTH',
  'EPISODE_ID',
  'STATE',
  'STATE_FIPS',
  'CZ_NAME',   # County name
  'CZ_TYPE',
  'CZ_FIPS',
  'EVENT_TYPE'
)
# Limit the dataframe to the selected columns
limited_stormdata <- storm_data %>% select(all_of(selected_columns))

# Display the first few rows of the limited dataframe
head(limited_stormdata)

```
# Arrange the dataframe by the state name

``` {r arrange-data_by_state, echo=FALSE}

# Arrange the dataframe by the state name
arranged_stormdata <- limited_stormdata %>% arrange(STATE)

# Display the first few rows of the arranged dataframe
head(arranged_stormdata)
```

## Arrange the stormdata by converting state and county name to Titlecase
``` {r Changestate/countyname-titlecase, echo=FALSE}
arranged_stormdata <- arranged_stormdata %>%
  mutate(
    STATE = str_to_title(STATE),
    CZ_NAME = str_to_title(CZ_NAME)
  )

# Display the first few rows of the modified data
head(arranged_stormdata)
rows_arranged = nrow(arranged_stormdata)
```

``` {r filter, echo=FALSE}
# Filter the events to include only those with CZ_TYPE of 'C'
filtered_storm_events <- arranged_stormdata %>%
  filter(CZ_TYPE == "C") %>%
  select(-CZ_TYPE)

rows_filtered = nrow(filtered_storm_events)
```
## Filter the storm events data to include only events listed by county FIPS.
The filtered data frame has `r rows_filtered` rows.

``` {r filter-display, echo=FALSE}
# Display the first few rows of the filtered dataframe
head(filtered_storm_events)
```

##  Unite the state and county FIPS into one FIPS column
``` {r padding, echo=FALSE}

# Pad the state and county FIPS with a "0" at the beginning
filtered_storm_events <- filtered_storm_events %>%
  mutate(
    STATE_FIPS = str_pad(STATE_FIPS, width = 3, side = "left", pad = "0"),
    CZ_FIPS = str_pad(CZ_FIPS, width = 3, side = "left", pad = "0")
  )

# Unite the state and county FIPS into one FIPS column
filtered_storm_events <- filtered_storm_events %>%
  unite("FIPS", STATE_FIPS, CZ_FIPS, sep = "")

# Display the first few rows
head(filtered_storm_events)
```

## Display the first few rows of the dataframe with lower case column names
```{r change_column_names_lowercase, echo=FALSE}
# Change all column names to lower case
filtered_storm_events <- filtered_storm_events %>%
  rename_all(tolower)

# Display the first few rows of the dataframe with lower case column names
head(filtered_storm_events)
```
## US State name, Area and Region
```{r US-states, echo=FALSE}
# Load the data on U.S. states
data("state")

# Create a dataframe with state name, area, and region
US_states <- data.frame(
  US_State = state.name,
  Area = state.area,
  Region = state.region
)

# Display the first few rows of the states dataframe
head(US_states)

```
## The number of storm events per state
```{r number-of-events-per-state, echo=FALSE}
# Create a dataframe with the number of events per state
events_per_state <- filtered_storm_events %>%
  group_by(state) %>%
  summarise(number_of_events = n())

# Display the first few rows of the events per state dataframe
head(events_per_state)
```
## Integrate state-level information, including area and region, from built-in R datasets.

```{r merge-data,echo = FALSE}

# Merge the events_per_state dataframe with the states_data dataframe
merged_data <- merge(events_per_state, US_states, by.x = "state", by.y = "US_State")

# Display the first few rows of the merged dataframe
kable(merged_data)
```

### Visualization

This Scatter plot is to visualize the relationship between land area and the number of storm events, with points color-coded by region.

```{r plotting, echo=FALSE}
# Create the scatter plot
ggplot(merged_data, aes(x = Area, y = number_of_events, color = Region)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "Northeast" = "red",
    "South" = "green",
    "North Central" = "blue",
    "West" = "purple"
  )) +
  labs(
    title = "Number of Storm Events in 1986 vs Land Area",
    x = "Land Area (square miles)",
    y = "Number of Storm Events"
  ) +
  theme_minimal()
```

